### 2017-01-28 introduction.tex
Corrections made to Introduction

Plasmas are fluid masses of charged particles, often formed when neutral atoms and molecules in a gas are ionized by external energy.  Instances of plasma include everything from florescent lightbulbs and flames to more exotic examples such as nuclear fusion reactor experiments, most stars, nebula, and the interstellar medium.  In fact, about 99\% of matter in the known universe exists in the plasma state.  Of particular interest to this thesis is the plasma in Earth's ionosphere.

Ionospheric plasma is partially ionized, meaning only a fraction of the particles are ionized and the rest are neutral.  Considering the dynamics of plasmas therefore requires not only collisions between particles to be considered but also charge-charge interactions and external and internal electric and magnetic fields.  These dynamics can generate waves in the plasma, which cause density perturbations.  Additionally, variations in temperature and electric and magnetic field strength can occur on a variety of scales.  This thesis focuses on mechanisms by which density variations in ionospheric plasma occur, often referred to as plasma structures.

\section{The Solar-Terrestrial Environment}
The solar terrestrial environment begins with the Sun in the center of the solar system.  The Sun is composed of highly energized plasma that is gravitationally bound together.  The surface of the Sun is highly dynamic, populated with sunspots, which are dark, relatively cool regions with intense magnetic fields, as well as coronal holes, which are low density areas characterized with a continuous outflow of plasma.  The intense magnetic fields associated with sunspots can break down releasing a large amount of energy and plasma, known as a solar flare.  Planet-sized masses of plasma known as coronal mass ejections (CMEs) can be ejected from the surface, which can be large enough to maintain an internal magnetic field as they travel through the solar system.  The frequency of sunspots and large outbursts of plasma changes with the 11-year long solar cycle.  A large number of sunspots, high occurrence of fast, dense plasma outbursts, and a highly variably magnetic field indicate solar maximum.  Conversely, solar minimum is characterized by a low sunspot number and slow and relatively steady outflow.  This thesis examines factors that control irregularity production in the Earth's ionosphere, which can be strongly influenced by solar activity.

The Sun interacts with the rest of the solar system through the solar wind, a continuous outflow of plasma that caries the Sun's magnetic field with it.  This is shown in Figure \ref{fig:ste}.  The solar wind, represented by the green arrows, typically has a density around 10 cm\(^{-3}\) and is traveling away from the Sun at approximately 400 km/s at Earth's orbit.  However, like the Sun, the solar wind is highly dynamic and these parameters can change drastically over a period of minutes.  CMEs, for instance, create a substantial increase in density and are often preceded by a shock, a region of hot, compressed plasma caused by the CME moving at supersonic speeds.  In addition to the outflow of plasma, the solar wind also consists of the Sun's magnetic field extended into the outer limits of the solar system, called the interplanetary magnetic field (IMF), dark blue lines in Figure \ref{fig:ste}.  Because the Sun has a short rotation period (about 33 days), the IMF gets twisted into a spiral, commonly called the Parker Spiral \citep{Parker1958}.  When the IMF reaches the Earth, the magnetic field is typically oriented at a \(45\deg\) inclination relative to the ecliptic, but this is highly variable depending on solar wind conditions.

\begin{figure}
	\includegraphics[width=\textwidth,angle=0]{ste.pdf}
	\caption[Solar-Terrestrial Environment]{Diagram of the solar-terrestrial environment.  The solar wind, green arrows, contains both a steady stream of charged particles and the IMF, dark blue.  The red region represents the magnetosheath.  The manetopause is shown by the black dashed line.  Closed magnetic field lines from the earth are shown in grey while open field lines are shown in light blue.  The ionosphere is shown in light green above the Earth's surface.}
	\label{fig:ste}
\end{figure}

The IMF interacts with the Earth's own magnetic field, and this interaction creates the magnetosphere.  Without the IMF, the magnetosphere would be approximately a dipole field, but dynamic pressure from the solar wind causes the Sun facing side to be compressed to 6--10 Earth radii and the rear side to be stretched into an extended tail (hundreds of Earth radii).  Because the solar wind is typically a supersonic flow, a bow shock forms on the dayside of the magnetosphere.  The heated and compressed solar wind plasma created by the bow shock is known as the magnetosheath, the red region in Figure \ref{fig:ste}.  The magnetopause, black dashed line, is the actual boundary between solar wind plasma in the magnetosheath and magnetospheric plasma.  Magnetic field lines that are part of the closed magnetosphere are shown in grey in Figure \ref{fig:ste}.  However, the magnetosphere is not fully closed.  At the magnetopause, magnetic reconnection can occur between the IMF and the Earth's magnetic field creating open field lines, light blue in Figure \ref{fig:ste}.  This reconnection occurs on last closed field line intersecting the magnetopause (highlighted in yellow in Figure \ref{fig:ste}).  Magnetic field lines originating from the Earth are then directly connected to the Sun through the solar wind, which consequently allows highly energized solar particles to enter the magnetosphere.  Open field lines tend to occur around the Earth's polar caps, making these regions particularly interesting to study due to very complex behavior and coupled interactions between the Earth and the Sun.

Closer to the Earth's surface, the magnetosphere also interacts with the Earth's ionosphere, a layer of partially ionized plasma in the upper atmosphere, light green in Figure \ref{fig:ste}.  The interaction between charged particles in the ionosphere with magnetic field lines couples the ionosphere, magnetosphere, and solar wind to create non-trivial dynamics.  Additionally,  neutral winds in the thermosphere can influence the ionosphere through viscous interactions.  Overall, the charged particle interactions in the ionosphere create a highly complex system, which is further complicated by coupling with both the thermosphere from below and the magnetosphere and solar wind from above.

\section{The Earth's Ionosphere}
\label{sec:ionosphere}
The Earth's ionosphere is a region of the upper atmosphere that ranges approximately between 50--1000 km in altitude where neutral gases have been partially ionized, resulting in free electrons and ions mixed with neutral particles.  Two competing processes occur in the ionosphere to create a peak in electron density.  Solar illumination and particle precipitation ionize neutrals, and the density of ions increase as altitude decreases because there are more neutral molecules available to ionize.  However, below a certain point, the concentration of ions becomes large enough that recombination into neutral particles becomes a substantial factor, reducing the ion density.  The exact altitude where this density peak occurs is variable with time of day, location, season, and solar cycle.  A profile of how electron density changes with altitude is shown in Figure \ref{fig:densprofile}, found from the International Reference Ionosphere (IRI) model run on January 1, 2007 at \(75\deg\) N, \(0\deg\) E, geographic (the IRI model is discussed further in Section \ref{sec:iri}).

\begin{figure}
	\includegraphics[width=\textwidth]{densprofile.pdf}
	\caption[Ionosphere density profile]{Altitude profile of electron density throughout the ionosphere found from the IRI model.  Typical ranges of the \(D\), \(E\), and \(F\) regions are identified.}
	\label{fig:densprofile}
\end{figure}

@@

\subsection{Ionospheric Regions}
\label{sec:ionosphere_regions}
The ionosphere is typically divided into three regions, Figure \ref{fig:densprofile}.  The D region is located between 80 and 90 km in altitude and typically only exists in the daytime when the ionosphere is sunlit.  The E region typically exists between 90 and 130 km with a density peak between 105 and 110 km, depending on factors such as time of day and season.  Above 130 km is considered to be the F region, which has a density peak around 250 km, Figure \ref{fig:densprofile} \citet{Luhmann1995}.  The F region can actually have two peaks under some daytime conditions, notated as the F1 peak and the F2 peak.  The peak heights and densities of each of these regions is highly variable both diurnally and seasonally.  In Figure \ref{fig:densprofile}, the daytime profile is shown with the solid line next to a nighttime profile with a dashed line.  Note that night electron densities are lower at all altitudes due to the lack of photoionization during nighttimes.  In addition, the F-region peak moves to higher altitudes at night.  The focus of this thesis will be the F region.

Because the concentrations of ions and neutral particles changes with altitude, each of these regions have distinct plasma physical properties that make them unique.  The main issue is whether the the large-scale motion of a particle is controlled more by the magnetic field (the particle is magnetized) or by collisions with other particles (the particle is collisional).  This can be determined by comparing the collision frequency, \(\nu_\alpha\), to the gyrofrequency, \(\Omega_\alpha\), of a particular species, where \(\alpha\) corresponds to either electrons, \(e\), or ions, \(i\).  If \(\Omega_\alpha \gg \nu_\alpha\), the particle completes many gyrations between each collision, so its motion is determined mostly by the magnetic field and it is magnetized.  If \(\nu_\alpha \gg \Omega_\alpha\), the particle collides with other particles much more often than it complete a gyration, so it is considered collisional.  The gyrofrequency is dependent on the mass and charge of the particle and the strength of the magnetic field, \(\Omega_\alpha = q_\alpha B/m_\alpha\), so is roughly constant through the ionosphere because the magnetic field strength does not change much throughout these altitudes.  The collision frequency depends on the density of a species relative to the density of all other species in the plasma and can be calculated from a series of standard expressions \citep{Schunk1980,Schunk2009}.  In the E region, \(\Omega_e \gg \nu_e\), so electrons are magnetized.  However, the motion on ions is dominated by collisions between ions and neutrals (\(\nu_i > \Omega_i\)), so ions in the \(E\) region are considered collisional.  In the \(F\) region where the neutral density is much lower, both ions and electrons are magnetized, \(\Omega_\alpha \gg \nu_\alpha\).  These differences in how charged particles move impact the conductivity and plasma waves development, which is further discussed in Section \ref{sec:lit_theory}.

@@

This thesis is concerned with the polar ionosphere, which is a particularly interesting and dynamic region due to the presence of open field lines to the solar wind.  The main focus of this work will be plasma density perturbations, often called irregularities.  Density irregularities in the polar cap can occur on scales ranging from thousands of kilometers to less than a centimeter \citep{Tsunoda1988}, and a summary of some of these irregularities is presented in Figure \ref{fig:polarirreg}.

Large-scale plasma irregularities are typically considered to be any density structure on the scale of greater than 30 km \citep{Kelley2009}.  Some common examples are polar patches, polar holes, and sun-aligned arcs.  Polar patches are large density enhancements (usually at least twice the background plasma density) that travel across the polar cap with the background convection, Figure \ref{fig:polarirreg}a \citep{Weber1984,Valladares1994}.  Polar holes are plasma density depletions that typically occur in the \(F\) region slightly poleward of the auroral oval \citep{Benson2001}.  Sun-aligned arcs are large density forms that stretch across the polar cap along the sun-earth line.  They are very narrow and often characterized by large and complex velocity shears on either side of the arc \citep{Valladares1991}.  Large scale plasma structures and the density gradients that surround them often serve as a platform for smaller-scale structures to form as well.  Polar patches, for instance, are known to form large finger-like structures along their trailing edges, Figure \ref{fig:polarirreg}b \citep{Gondarenko2004b,Hosokawa2016}.

\begin{figure}
	\includegraphics[width=0.5\textwidth]{irregularity.pdf}
	\caption[Plasma irregularities in the polar ionosphere]{Plasma density irregularities occur in the polar cap at different scale sizes.  Polar patches, orange, move with the background ionospheric convection, indicated by red contours (a).  Smaller-scale structures often move develop along the edges of polar patches where strong density gradients exist (b).  Intermediate scale irregularities are primarily responsible for disrupting radio signals traveling through the ionosphere (c).  Field-aligned irregularities, FAIs, are density perturbations that form along the magnetic field lines, light blue (d).}
	\label{fig:polarirreg}
\end{figure}

Intermediate-scale plasma irregularities are those ranges from 100 m -- 30 km \citep{Kelley2009}.  These are also referred to as scintillation-causing structures and are responsible for many of the negative space weather effects that are observed on Earth.  One important example of a space weather effect is when the ionosphere changes the phase and amplitude of the original signal, called radio scintillation, which affects the ability of satellites to communicate with ground receivers.  This can cause communication blackouts or introduce large errors into GPS/GNSS (Global Positioning System/Global Navigation Satellite System) calculations, affecting navigation.

Small-scale structuring is generally considered to be all irregularities less than 100 m \citep{Kelley2009}.  Although small-scale irregularities are not responsible for scintillation that directly impact GPS/GNSS performance, they can be the easiest to study because of large data sets collected from high frequency ionospheric radars globally for several decades (discussed further in Section \ref{sec:superdarn}).  Because electrons move far more easily along magnetic field lines than across them, density perturbations tend to be aligned with the magnetic field and are often known as field-aligned irregularities (FAI), Figure \ref{fig:polarirreg}d.  The irregularity classifications presented here are rough categories only and a high degree of coupling between different scales occurs, which makes plasma structures of all scales (large, intermediate, and small) important for plasma structuring processes.  For instance, large-scale structures such as polar patches may very well have internal intermediate- and/or small-scale structuring.  A turbulent cascade is often assumed to connect these different scales, but because various instruments or techniques usually only observe structuring on a particular scale, it is difficult to find direct evidence of this or how exactly it occurs \citep{Kintner1985,Tsunoda1985}.  A description of some of the particular instruments and models that are used in this thesis follows.

@@

There are three measurements typically made by CSR systems: backscatter power, line-of-sight (LoS) velocity, and spectral width.  Backscatter power is the power of the returned signal that the radar's receivers detect.  Typically, a threshold is selected for the minimum power acceptable for a return to be considered an actual signal instead of background noise.  LoS velocity is the component of the total plasma drift velocity that is measured along the transmitted signal direction.  Because radars measure velocity through the doppler shift of backscatter, only the component of the velocity vector can be measured.  Spectral width represents the width of the doppler power spectrum and can determine characteristics of irregularities \citep{Greenwald1985}.

CSR systems can operate by transmitting pulses instead of continuously.  This allows the distance between the radar and the target to be found (based on the time difference between when the pulse is transmitted and received).  The size of the range gate is determined by the length of a radar pulse, Equation \ref{eqn:range_gate}.
\begin{equation}
	\label{eqn:range_gate}
	\Delta r = \frac{c\Delta t}{2}
\end{equation}.
Here, \(\Delta r\) is the size of a range gate, \(\Delta t\) is the pulse length, and \(c\) is the speed of light.  Longer pulses result in larger range gates, resulting in poorer spatial resolution.

When the radar transmits continuously, backscatter will be received, but because it is impossible to know when that signal was transmitted, the difference in time between transmission and receiving cannot be used to calculate how far away the backscatter volume is.  Conversely, if the radar only transmits a pulse of a certain length of a certain length and then ``listens'' for its return, the time between transmission and return is known and the distance the pulse traveled can be calculated and therefore the location of the backscattering volume \citep{Farley1972,Greenwald1983}.  However, using this simple single-pulse method another pulse cannot be transmitted until the first pulse returns, limiting the temporal resolution that can be achieved.  This can be improved by using a multipulse scheme, which will be discussed below \citep{Farley1972,Greenwald1983,Greenwald1985}.

A short time between pulse returns is necessary to allow Doppler velocities to be calculated, particularly at long ranges.  The maximum Doppler shifted frequency that can be observed by a radar is the Nyquist frequency, \(f_n\), which is equivalent to half the sampling frequency, \(f_s\), Equation \ref{eqn:nyquist}
\begin{equation}	
	\label{eqn:nyquist}
	f_n = \frac{f_s}{2}
\end{equation}.
Radars operating in single-pulse mode must have a very low sampling frequency, as described above, particularly at long ranges where the pulse must travel a long distance.  This severely limits the Doppler shift that can be measured, which places an upper limit on the LoS Doppler velocity that can be calculated with this technique.  For an ionospheric radar observing a structure 2000 km away, it would take about 13 ms for a signal traveling at the speed of light to travel to the the structure and back to the radar, which corresponds to a sampling frequency of \(f_s = 75\) Hz.  The Nyquist frequency is then \(f_n = 37.5\) Hz, which corresponds to a maximum measurable Doppler velocity of \(\sim550\) m/s.  Flows in the ionosphere have been known to exceed 1000--2000 m/s, so single pulse sampling is insufficient for these purposes.

Instead of waiting for each individual signal to be received before transmitting the next, ionospheric radars typically employ a multipulse mode. In a multipulse mode, the radar transmits a series of pules with different time intervals or lags between them.  This can introduce complications when backscatter from different pulses at different ranges is received by the radar at the same time, referred to as cross-range interference, but these effects can be mitigated using correlation techniques \citep{Farley1972}.  The smallest lag between two pulses is known as the multi-pulse increment, \(\tau\).  All other lags are integer multiples of \(\tau\), which allows the calculation of the corresponding lags of the complex autocorrelation function (ACF).  The ACF is used to find the spectral characteristics of the backscatter, from which the LoS velocity, spectral width, and backscatter power can be obtained.  Overall, multi-pulse techniques are generally considered far more appropriate for ionospheric studies than single pulse techniques \citep{Farley1972,Greenwald1983,Greenwald1985,Barthes1998,Ponomarenko2006}.

The first observations of the ionosphere by CSRs were made in the 1930s \citep{Eckersley1937,Harang1938}.  Over the next several decades, numerous used ground-based radars to advance our understanding of the ionosphere and radio aurora \citep{Hultqvist1964,Leadabrand1965,Unwin1972,Sahr1996}.  In the 1970s and 1980s, the Scandanavian Twin Auroral Radar Experiment (STARE) produced the first continuous large data-set, similar to how modern CSR systems are run.  STARE consisted of two very high frequency (VHF) radars with overlapping FoVs that were designed to measure FAIs in the \(E\) region of the ionosphere \citep{Greenwald1997}.  The advantage of two radars with overlapping FoVs was the ability to observe the same structures from two different directions.  Because each radar can only measure the LoS velocity of a structure, two simultaneous observations from different directions allows two different velocity vector components to be found, and hence the total velocity vector can be calculated.  This technique is still commonly used with CSR system.

>>>>

Plasmas are fluid masses of charged particles, often formed when neutral atoms and molecules in a gas are ionized by external energy.  Instances of plasma include everything from florescent lightbulbs and flames to more exotic examples such as nuclear fusion reactor experiments, most stars, nebulas, and the interstellar medium.  In fact, about 99\% of visible matter in the known universe exists in the plasma state.  Of particular interest to this thesis is the plasma in Earth's ionosphere.

Ionospheric plasma is partially ionized, meaning only a fraction of the particles are ionized and the rest are neutral.  Considering the dynamics of plasmas therefore requires not only collisions between particles to be considered but also charge-charge interactions and external and internal electric and magnetic fields.  These dynamics can generate waves in the plasma, which cause density perturbations.  Similarly, variations in plasma temperature and electric and magnetic field strength can occur on a variety of scales.  This thesis focuses on mechanisms by which density variations in ionospheric plasma occur, often referred to as plasma structures.

\section{The Solar-Terrestrial Environment}
The solar-terrestrial environment begins with the Sun in the center of the solar system.  The Sun is composed of highly energized plasma that is gravitationally bound together.  The surface of the Sun is highly dynamic, populated with sunspots, which are dark and relatively cool regions with intense magnetic fields, as well as coronal holes, which are low-density areas characterized with a continuous outflow of plasma.  The intense magnetic fields associated with sunspots can break down releasing a large amount of energy and plasma, known as a solar flare.  Planet-sized masses of plasma known as coronal mass ejections (CMEs) can be ejected from the surface, which can be large enough to maintain an internal magnetic field as they travel through the solar system.  The frequency of sunspots and large outbursts of plasma changes with the 11-year-long solar cycle.  A large number of sunspots, high occurrence of fast and dense plasma outbursts, and a highly variable magnetic field indicate solar maximum.  Conversely, solar minimum is characterized by a low sunspot number and slow and relatively steady outflow.  This thesis examines factors that control irregularity production in the Earth's ionosphere, which can be strongly influenced by solar activity.

The Sun interacts with the rest of the solar system through the solar wind, a continuous outflow of plasma that carries the Sun's magnetic field with it.  This is shown in Figure \ref{fig:ste}.  The solar wind, represented by the green arrows in Figure \ref{fig:ste}, typically has a density around 10 cm\(^{-3}\) and is traveling away from the Sun at approximately 400 km/s at the Earth's orbit.  However, like the Sun, the solar wind is highly dynamic and these parameters can change drastically over a period of minutes.  CMEs, for instance, create a substantial increase in density and are often preceded by a shock, a region of hot, compressed plasma caused by the CME moving at supersonic speeds.  In addition to the outflow of plasma, the solar wind also contains the Sun's magnetic field extended into the outer limits of the solar system, called the interplanetary magnetic field (IMF), dark blue lines in Figure \ref{fig:ste}.  Because the Sun has a short rotation period (about 33 days), the IMF gets twisted into a spiral, commonly called the Parker Spiral \citep{Parker1958}.  When the IMF reaches the Earth, the magnetic field is typically oriented at a \(45\deg\) inclination relative to the ecliptic, but this is highly variable depending on solar wind conditions.

\begin{figure}
	\includegraphics[width=\textwidth,angle=0]{ste.pdf}
	\caption[Solar-terrestrial environment]{Diagram of the solar-terrestrial environment.  The solar wind, green arrows, contains both a steady stream of charged particles and the interplanetary magnetic field, dark blue.  The red region represents the magnetosheath, the shock layer formed due to the supersonic plasma outflow from the Sun.  The magnetopause encloses the magnetosphere dominated by the geomagnetic field and is shown by the black dashed line.  Closed magnetic field lines from the Earth are shown in grey, while open field lines are shown in light blue.  The ionosphere is shown in yellow above the Earth's surface.}
	\label{fig:ste}
\end{figure}

The IMF interacts with the Earth's own magnetic field, and this interaction creates the magnetosphere.  Without the IMF, the magnetosphere would be approximately in a dipole shape, but dynamic pressure from the solar wind causes the Sun-facing side to be compressed to 6--10 Earth radii and the opposite side to be stretched into an extended tail (hundreds of Earth radii).  Because the solar wind is typically a supersonic flow, a bow shock forms on the dayside of the magnetosphere.  The heated and compressed solar wind plasma created by the bow shock is known as the magnetosheath, the red region in Figure \ref{fig:ste}.  The magnetopause, black dashed line, is the actual boundary between solar wind plasma in the magnetosheath and magnetospheric plasma.  Magnetic field lines that are part of the closed magnetosphere are shown in grey in Figure \ref{fig:ste}.  However, the magnetosphere is not fully closed.  At the magnetopause, magnetic reconnection can occur between the IMF and the Earth's magnetic field creating open field lines, light blue in Figure \ref{fig:ste}.  This reconnection occurs on last closed field line intersecting the magnetopause (highlighted in yellow in Figure \ref{fig:ste}).  Magnetic field lines originating from the Earth are then directly connected to the Sun through the solar wind, which consequently allows highly energized solar particles to enter the magnetosphere.  Open field lines tend to occur around the Earth's polar caps, making these regions particularly interesting to study due to very complex behavior and coupled interactions between the Earth and the Sun.

Closer to the Earth's surface, the magnetosphere also interacts with the Earth's ionosphere, a layer of partially ionized plasma in the upper atmosphere, yellow in Figure \ref{fig:ste}.  The interaction between charged particles in the ionosphere with magnetic field lines couples the ionosphere, magnetosphere, and solar wind to create non-trivial dynamics.  Additionally,  neutral winds in the thermosphere can influence the ionosphere through viscous interactions.  Overall, the charged particle interactions in the ionosphere create a highly complex system, which is further complicated by coupling with both the thermosphere from below and the magnetosphere and solar wind from above.

\section{The Earth's Ionosphere}
\label{sec:ionosphere}
The focus of this body of work is the Earth's ionosphere, a region of the upper atmosphere that extends approximately between 50--1000 km in altitude where neutral gases have been partially ionized, resulting in free electrons and ions mixed with neutral particles.  Two competing processes occur in the ionosphere to create a peak in the electron density.  Solar illumination and particle precipitation ionize neutrals, and the density of ions increases as altitude decreases because there are more neutral molecules available to ionize.  However, below a certain point, the concentration of ions becomes large enough that recombination becomes a substantial factor, reducing the ion density.  The exact altitude where this density peak occurs varries with time of day, location, season, and solar cycle.  Representative model examples of electron density changes with altitude are shown in Figure \ref{fig:densprofile}, found from the International Reference Ionosphere (IRI) model run on January 1, 2007 at \(75\deg\) N, \(0\deg\) E, geographic (the IRI model is discussed further in Section \ref{sec:iri}).

\begin{figure}
	\includegraphics[width=\textwidth]{densprofile.pdf}
	\caption[Ionosphere density profile]{Altitude profiles of electron density throughout the ionosphere found from the IRI model.  Typical extents of the \(D\), \(E\), and \(F\) regions are identified.}
	\label{fig:densprofile}
\end{figure}

@@

\subsection{Ionospheric Regions}
\label{sec:ionosphere_regions}
The ionosphere is typically divided into three regions, Figure \ref{fig:densprofile}.  The \(D\) region is located between 80 and 90 km in altitude and typically only exists in the daytime when the ionosphere is sunlit.  The \(E\) region typically exists between 90 and 130 km with a density peak between 105 and 110 km, depending on factors such as time of day and season.  Above 130 km is considered to be the F region, which has a density peak around 250 km, Figure \ref{fig:densprofile} \citep{Kivelson1995}.  The \(F\) region can actually have two peaks under some daytime conditions, notated as the \(F1\) peak and the \(F2\) peak.  The peak heights and densities of each of these regions is highly variable both diurnally and seasonally.  In Figure \ref{fig:densprofile}, the daytime profile is shown with the solid line next to a nighttime profile shown with a dashed line.  Note that night electron densities are lower at all altitudes due to the lack of photoionization during the night time.  In addition, the \(F\)-region peak moves to higher altitudes at night.  The focus of this thesis is on the \(F\) region.

Because the concentrations of ions and neutral particles changes with altitude, each of these regions have distinct plasma physical properties that make them unique.  The main issue is whether the large-scale motion of a particle is controlled more by the magnetic field (the particle is magnetized) or by collisions with other particles (the particle is collisional).  This can be determined by comparing the collision frequency, \(\nu_\alpha\), to the gyrofrequency, \(\Omega_\alpha\), of a particular species, where \(\alpha\) refers to either electrons, \(e\), or ions, \(i\).  If \(\Omega_\alpha \gg \nu_\alpha\), the particle completes many gyrations between each collision, so its motion is determined mostly by the magnetic field and it is magnetized.  If \(\nu_\alpha \gg \Omega_\alpha\), the particle collides with other particles much more often than it complete a gyration, so it is considered collisional.  The gyrofrequency is dependent on the mass and charge of the particle and the strength of the magnetic field, \(\Omega_\alpha = q_\alpha B/m_\alpha\), so is roughly constant throughout the ionosphere because the magnetic field strength does not change much throughout these altitudes.  The collision frequency depends on the density of a species relative to the density of all other species in the plasma and can be calculated from a series of standard expressions \citep{Schunk1980,Schunk2009}.  In the E region, \(\Omega_e \gg \nu_e\), so electrons are magnetized.  However, the motion on ions is dominated by collisions between ions and neutrals (\(\nu_i > \Omega_i\)), so ions in the \(E\) region are considered collisional.  In the \(F\) region where the neutral density is much lower, both ions and electrons are magnetized, \(\Omega_\alpha \gg \nu_\alpha\).  These differences in how charged particles move impact the conductivity and plasma waves development, which is further discussed in Section \ref{sec:lit_instabilities}.

@@

This thesis is concerned with the polar ionosphere, which is a particularly interesting and dynamic region due to the presence of open field lines that are directly connected to the solar wind.  The main focus of this work is plasma density waves or structures, also known as irregularities.  Density irregularities in the polar cap can occur on scales ranging from thousands of kilometers to less than a centimeter \citep{Tsunoda1988}, and a summary of some of these irregularities is presented in Figure \ref{fig:polarirreg}.

Large-scale plasma irregularities are typically considered to be any density structure on the scale of greater than 30 km \citep{Kelley2009}.  Some common examples are polar patches, polar holes, and sun-aligned arcs.  Polar patches are large density enhancements (usually at least twice the background plasma density) that travel across the polar cap with the background convection, Figure \ref{fig:polarirreg}a \citep{Weber1984,Valladares1994}.  Polar holes are plasma density depletions that typically occur in the \(F\) region slightly poleward of the auroral oval \citep{Brinton1978}.  Sun-aligned arcs are large density forms that stretch across the polar cap along the Sun-Earth line.  They are very narrow and often characterized by large and complex velocity shears on either side of the arc \citep{Valladares1991}.  Large-scale plasma structures and the density gradients that surround them often serve as a platform for smaller-scale structures to form as well.  Polar patches, for instance, are known to form large finger-like structures along their trailing edges, Figure \ref{fig:polarirreg}b \citep{Gondarenko2004b,Hosokawa2016}.

\begin{figure}
	\includegraphics[width=0.5\textwidth]{irregularity.pdf}
	\caption[Plasma irregularities in the polar ionosphere]{Plasma density irregularities occur in the polar cap at different scale sizes.  Polar patches, orange in Figures \ref{fig:polarirreg}a and \ref{fig:polarirreg}b, move with the background ionospheric convection, indicated by red arrows.  Smaller-scale structures often develop along the edges of polar patches where strong density gradients exist, Figure \ref{fig:polarirreg}b.  Intermediate-scale irregularities are primarily responsible for disrupting radio signals traveling through the ionosphere, Figure \ref{fig:polarirreg}c.  Field-aligned irregularities, FAIs, are density perturbations that form along the magnetic field lines, Figure \ref{fig:polarirreg}d.}
	\label{fig:polarirreg}
\end{figure}

Intermediate-scale plasma irregularities are those with sizes of 100 m -- 30 km \citep{Kelley2009}.  These are also referred to as scintillation-causing structures and are responsible for many of the aderse space weather effects that are observed on Earth.  One important example of a space weather effect is when the ionosphere changes the phase and amplitude of the original signal, called radio scintillation, which affects the ability of satellites to communicate with ground receivers.  This can cause communication blackouts or introduce large errors into GPS/GNSS (Global Positioning System/Global Navigation Satellite System) calculations, affecting navigation.

Small-scale structuring is generally considered to be all irregularities less than 100 m \citep{Kelley2009}.  Although small-scale irregularities are not responsible for scintillation that directly impact GPS/GNSS performance, they can be the easiest to study because of large datasets collected with a large number of ionospheric radars around the world that have been in operation for many decades (discussed further in Section \ref{sec:superdarn}).  Because electrons move far more easily along magnetic field lines than across them, density perturbations tend to be aligned with the magnetic field and are often known as field-aligned irregularities (FAI), Figure \ref{fig:polarirreg}d.  The irregularity classifications presented here are rough categories only and a high degree of coupling between different scales occurs, which makes plasma structures of all scales (large, intermediate, and small) important for plasma structuring processes.  For instance, large-scale structures such as polar patches may very well have internal intermediate- and/or small-scale structuring.  A turbulent cascade is often assumed to connect these different scales, but because various instruments or techniques usually only observe structuring on a particular scale, it is difficult to find direct evidence of this or how exactly it occurs \citep{Kintner1985,Tsunoda1985}.  A description of some of the particular instruments and models that are used in this thesis follows.

@@

There are three measurements typically made by CSR systems: backscatter power, line-of-sight (LoS) velocity, and spectral width.  Backscatter power is the power of the returned signal that the radar's receivers detect.  Typically, a threshold is selected for the minimum power acceptable for a return to be considered an actual signal instead of background noise.  LoS velocity is the component of the total plasma irregularity velocity that is measured along the transmitted signal direction.  Because radars measure velocity through the Doppler shift of backscatter, only a component of the velocity vector can be measured.  Spectral width represents the width of the Doppler power spectrum and can be used to infer other characteristics of irregularities, such as their lifetime \citep{Greenwald1985}.

CSR systems can operate by transmitting pulses instead of continuously.  This allows the distance between the radar and the target to be found (based on the time difference between when the pulse is transmitted and received).  The size of the range gate is determined by the length of a radar pulse, Equation \ref{eqn:range_gate}
\begin{equation}
	\label{eqn:range_gate}
	\Delta r = \frac{c\Delta t}{2}.
\end{equation}
Here, \(\Delta r\) is the size of a range gate, \(\Delta t\) is the pulse length, and \(c\) is the speed of light.  Longer pulses result in larger range gates, resulting in poorer spatial resolution.

When the radar transmits continuously, backscatter will be received, but because it is impossible to know when that signal was transmitted, the difference in time between transmission and receiving cannot be used to calculate how far away the backscatter volume is.  Conversely, if the radar only transmits a pulse of a certain length of a certain length and then ``listens'' for its return, the time between transmission and return is known and the distance the pulse traveled can be calculated and therefore the location of the backscattering volume can be found \citep{Farley1972,Greenwald1983}.  However, using this simple single-pulse method another pulse cannot be transmitted until the first pulse returns, limiting the temporal resolution that can be achieved.  This can be improved by using a multipulse scheme, which will be discussed below \citep{Farley1972,Greenwald1983,Greenwald1985}.

A short time between pulse returns is necessary to allow Doppler velocities to be calculated, particularly at long ranges.  The maximum Doppler shifted frequency that can be observed by a radar is the Nyquist frequency, \(f_n\), which is equivalent to half the sampling frequency, \(f_s\), Equation \ref{eqn:nyquist}
\begin{equation}	
	\label{eqn:nyquist}
	f_n = \frac{f_s}{2}.
\end{equation}
Radars operating in single-pulse mode must have a very low sampling frequency, as described above, particularly at long ranges where the pulse must travel a long distance.  This severely limits the Doppler shift that can be measured, which places an upper limit on the LoS velocity that can be calculated with this technique.  For an ionospheric radar observing a structure 2000 km away, it would take about 13 ms for a signal traveling at the speed of light to travel to the structure and back to the radar, which corresponds to a sampling frequency of \(f_s = 75\) Hz.  The Nyquist frequency is then \(f_n = 37.5\) Hz, which corresponds to a maximum measurable Doppler velocity of \(\sim550\) m/s.  Flows in the ionosphere have been known to exceed 1000--2000 m/s, so single pulse sampling is insufficient for these purposes.

Instead of waiting for each individual signal to be received before transmitting the next, ionospheric radars typically employ a multipulse mode. In a multipulse mode, the radar transmits a series of pulses with different time intervals or lags between them.  This can introduce complications when backscatter from different pulses at different ranges is received by the radar at the same time, referred to as cross-range interference, but these effects can be mitigated using correlation techniques \citep{Farley1972}.  The smallest lag between two pulses is known as the multi-pulse increment, \(\tau\).  All other lags are integer multiples of \(\tau\), which allows the calculation of the corresponding lags of the complex autocorrelation function (ACF).  The ACF is used to find the spectral characteristics of the backscatter, from which the LoS velocity, spectral width, and backscatter power can be obtained.  Overall, multi-pulse techniques are generally considered far more appropriate for ionospheric studies than single pulse techniques \citep{Farley1972,Greenwald1983,Greenwald1985,Barthes1998,Ponomarenko2006}.

The first observations of the ionosphere by CSRs were made in the 1930s \citep{Eckersley1937,Harang1938}.  Over the next several decades, numerous ground-based radars have been used to advance our understanding of the ionosphere and radio aurora \citep{Hultqvist1964,Leadabrand1965,Unwin1972,Sahr1996}.  In the 1970s and 1980s, the Scandanavian Twin Auroral Radar Experiment (STARE) produced the first continuous large data-set, similar to how modern CSR systems are run.  STARE consisted of two very high frequency (VHF) radars with overlapping fields-of-view (FoV) that were designed to measure FAIs in the \(E\) region of the ionosphere \citep{Greenwald1997}.  The advantage of two radars with overlapping FoVs was the ability to observe the same structures from two different directions.  Because each radar can only measure the LoS velocity of a structure, two simultaneous observations from different directions allows two different velocity vector components to be found, and hence the total velocity vector can be calculated.  This technique is still commonly used with CSR system.

\subsubsection{Super Dual Auroral Radar Network (SuperDARN)}
\label{sec:superdarn}
The primary instruments used within this body of work are radars within the Super Dual Auroral Radar Network (SuperDARN), a global network of high frequency (HF) CSRs that was designed to observe small-scale plasma structures in the ionosphere and map the global plasma convection.  The network currently consists of about 35 operational radars between the northern and southern hemispheres distributed at mid-, high-, and polar latitudes, Figure \ref{fig:superdarnmap}.  After the STARE experiment \citep{Greenwald1978}, the first HF ionospheric radar was built in Goose Bay, Canada in 1983, which would become the first of the SuperDARN radars \citep{Greenwald1985}.
